---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 7,
  out.width = "100%",
  dpi = 144
)

has_rayimage = requireNamespace("rayimage", quietly = TRUE)
has_rayrender = requireNamespace("rayrender", quietly = TRUE)
```

# libopenexr

<!-- badges: start -->
[![R-CMD-check](https://github.com/tylermorganwall/libopenexr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylermorganwall/libopenexr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`libopenexr` provides OpenEXR headers and static libraries for R packages
that need high-dynamic-range EXR image support. It also includes a small R
API for reading and writing simple RGBA EXR files.

The package is primarily intended for package authors. The R functions are
useful for quick tests and examples, but the main purpose of this package is
to make OpenEXR available to downstream packages without requiring users to
install OpenEXR and Imath manually.

## Installation

```{r install, eval=FALSE}
# once released on CRAN
install.packages("libopenexr")

# development version
remotes::install_github("tylermorganwall/libopenexr")
```

## Basic R Usage

`read_exr()` reads RGBA channels into numeric matrices. `write_exr()` writes
numeric RGBA matrices as a 32-bit float EXR file.

```{r basic-exr}
library(libopenexr)

exr_file = tempfile(fileext = ".exr")

write_exr(
  exr_file,
  r = widecolorgamut[, , 1],
  g = widecolorgamut[, , 2],
  b = widecolorgamut[, , 3],
  a = widecolorgamut[, , 4]
)

exr = read_exr(exr_file)
str(exr[c("width", "height")])

# Increase red-channel exposure and write a second EXR.
edited_file = tempfile(fileext = ".exr")
write_exr(
  edited_file,
  r = exr$r * 2,
  g = exr$g,
  b = exr$b,
  a = exr$a
)
```

```{r basic-exr-preview, eval=has_rayimage, message=FALSE, warning=FALSE}
rayimage::ray_read_image(edited_file) |>
  rayimage::render_tonemap(method = "hbd") |>
  rayimage::plot_image()
```

For image processing workflows, `rayimage` uses `libopenexr` internally when
reading or writing `.exr` files:

```{r rayimage-exr, eval=has_rayimage, message=FALSE, warning=FALSE}
library(rayimage)

input = tempfile(fileext = ".exr")
output = tempfile(fileext = ".exr")

write_exr(
  input,
  r = widecolorgamut[, , 1],
  g = widecolorgamut[, , 2],
  b = widecolorgamut[, , 3],
  a = widecolorgamut[, , 4]
)

processed = ray_read_image(input) |>
  render_exposure(exposure = -1) |>
  render_tonemap(method = "hbd")

ray_write_image(processed, output)
plot_image(processed)
```

`rayrender` can also save rendered output directly to EXR through
`rayimage::ray_write_image()`:

```{r rayrender-exr, eval=has_rayimage && has_rayrender, message=FALSE, warning=FALSE}
library(rayrender)

render_file = tempfile(fileext = ".exr")

scene = generate_cornell(lightwidth = 180, lightdepth = 180) |>
  add_object(
    sphere(
      x = 555 / 2,
      y = 120,
      z = 555 / 2,
      radius = 120,
      material = diffuse(color = "white")
    )
  )

render_scene(
  scene,
  filename = render_file,
  width = 320,
  height = 320,
  samples = 16,
  lookfrom = c(278, 278, -800),
  lookat = c(278, 278, 0),
  aperture = 0,
  fov = 40,
  tonemap = "raw",
  bloom = FALSE,
  plot_scene = FALSE,
  parallel = TRUE,
  progress = FALSE
)

rayimage::ray_read_image(render_file) |>
  rayimage::render_tonemap(method = "hbd") |>
  rayimage::plot_image()
```

## Using libopenexr in Another Package

Downstream packages should list both `libopenexr` and `libimath` in
`LinkingTo`:

```text
LinkingTo:
    libimath,
    libopenexr
SystemRequirements: OpenEXR (>= 4.0.0), Imath (>= 3.2.0)
```

The package ships static archives for convenience, but CRAN packages should
first try to use suitable system libraries. In practice, many build machines
have older OpenEXR and Imath versions, so your `configure` script should
check versions and fall back to the package-provided static libraries when
the system libraries are missing or too old.

I recommend using Kevin Ushey's R-based `configure.R` for configuring your package. It lets you
keep the configure logic in R, use the same logic on Unix and Windows, and
generate `src/Makevars` from `src/Makevars.in`. See the `tools/config.R`
helper used in `rayrender`, `skymodelr`, and this package for a complete
implementation.

Your top-level configure scripts can be simple wrappers:

```sh
#!/usr/bin/env sh
: "${R_HOME=`R RHOME`}"
"${R_HOME}/bin/Rscript" tools/config.R configure "$@"
```

```sh
#!/usr/bin/env sh
"${R_HOME}/bin${R_ARCH_BIN}/Rscript.exe" tools/config.R configure "$@"
```

### Configure Example

This is a compact `tools/config/configure.R` example. It checks for native
OpenEXR and Imath with `pkg-config`, requires compatible versions, and falls
back to the static archives installed by `libopenexr` and `libimath`.

```{r configure-example, eval=FALSE}
options(
  configure.common = FALSE,
  configure.platform = FALSE
)

is_windows = identical(.Platform$OS.type, "windows")
target_arch = Sys.info()[["machine"]]

openexr_min_version = "4.0.0"
imath_min_version = "3.2.0"
openexr_api = "4_0"
imath_api = "3_2"

quote_path = function(path) {
  shQuote(
    normalizePath(path, winslash = "/", mustWork = FALSE),
    type = if (is_windows) "cmd" else "sh"
  )
}

flag_path = function(flag, path) {
  paste0(flag, quote_path(path))
}

collapse_flags = function(flags) {
  flags = flags[nzchar(flags)]
  paste(flags, collapse = " ")
}

read_runtime_link_flags = function(path) {
  if (!file.exists(path)) {
    return("")
  }

  flags = unlist(strsplit(
    trimws(paste(readLines(path, warn = FALSE), collapse = " ")),
    "\\s+"
  ))
  collapse_flags(unique(flags))
}

pkgconfig = Sys.which("pkg-config")

pkg_config_exists = function(module, version) {
  if (!nzchar(pkgconfig)) {
    return(FALSE)
  }

  status = system2(
    pkgconfig,
    c("--exists", sprintf("%s >= %s", module, version)),
    stdout = FALSE,
    stderr = FALSE
  )

  identical(as.integer(status), 0L)
}

pkg_config_flags = function(args, module) {
  trimws(paste(
    system2(pkgconfig, c(args, module), stdout = TRUE),
    collapse = " "
  ))
}

use_system_openexr =
  pkg_config_exists("OpenEXR", openexr_min_version) &&
  pkg_config_exists("Imath", imath_min_version)

if (use_system_openexr) {
  message("*** configure: using system OpenEXR and Imath")

  OPENEXR_CPPFLAGS = collapse_flags(c(
    pkg_config_flags("--cflags", "OpenEXR"),
    pkg_config_flags("--cflags", "Imath")
  ))

  OPENEXR_LIBS = collapse_flags(c(
    pkg_config_flags("--libs", "OpenEXR"),
    pkg_config_flags("--libs", "Imath")
  ))
} else {
  message("*** configure: using libopenexr and libimath static libraries")

  openexr_include_root = system.file(
    "include",
    package = "libopenexr",
    mustWork = TRUE
  )
  openexr_include_dir = system.file(
    "include",
    "OpenEXR",
    package = "libopenexr",
    mustWork = TRUE
  )
  openexr_lib_arch = normalizePath(
    file.path(
      system.file("lib", package = "libopenexr", mustWork = TRUE),
      target_arch
    ),
    winslash = "/",
    mustWork = TRUE
  )

  imath_include_root = system.file(
    "include",
    package = "libimath",
    mustWork = TRUE
  )
  imath_include_dir = system.file(
    "include",
    "Imath",
    package = "libimath",
    mustWork = TRUE
  )
  imath_lib_arch = normalizePath(
    file.path(
      system.file("lib", package = "libimath", mustWork = TRUE),
      target_arch
    ),
    winslash = "/",
    mustWork = TRUE
  )

  openjph_static_lib = file.path(openexr_lib_arch, "libopenjph.a")
  openjph_link = if (file.exists(openjph_static_lib)) {
    quote_path(openjph_static_lib)
  } else {
    ""
  }

  openexr_runtime_flags = read_runtime_link_flags(file.path(
    openexr_lib_arch,
    "libopenexr-runtime-link-flags"
  ))
  imath_runtime_flags = read_runtime_link_flags(file.path(
    imath_lib_arch,
    "libimath-runtime-link-flags"
  ))

  OPENEXR_CPPFLAGS = collapse_flags(c(
    flag_path("-I", openexr_include_root),
    flag_path("-I", openexr_include_dir),
    flag_path("-I", imath_include_root),
    flag_path("-I", imath_include_dir)
  ))

  OPENEXR_LIBS = collapse_flags(c(
    flag_path("-L", openexr_lib_arch),
    sprintf("-lOpenEXRUtil-%s", openexr_api),
    sprintf("-lOpenEXR-%s", openexr_api),
    sprintf("-lOpenEXRCore-%s", openexr_api),
    openjph_link,
    sprintf("-lIlmThread-%s", openexr_api),
    sprintf("-lIex-%s", openexr_api),
    flag_path("-L", imath_lib_arch),
    sprintf("-lImath-%s", imath_api),
    openexr_runtime_flags,
    imath_runtime_flags
  ))
}

define(
  OPENEXR_CPPFLAGS = OPENEXR_CPPFLAGS,
  OPENEXR_LIBS = OPENEXR_LIBS
)

if (is_windows) {
  configure_file("src/Makevars.win.in")
} else {
  configure_file("src/Makevars.in")
}
```

### Makevars

Use the configured values in both `src/Makevars.in` and
`src/Makevars.win.in`:

```make
PKG_CPPFLAGS = @OPENEXR_CPPFLAGS@
PKG_LIBS = @OPENEXR_LIBS@ $(SAN_LIBS)

OBJECTS = my_exr_reader.o R_init_mypackage.o

$(SHLIB): $(OBJECTS)
```

For the bundled fallback, keep the static libraries in dependency order:

```make
-lOpenEXRUtil-4_0 \
-lOpenEXR-4_0 \
-lOpenEXRCore-4_0 \
libopenjph.a \
-lIlmThread-4_0 \
-lIex-4_0 \
-lImath-3_2
```

Static linkers resolve symbols from left to right, so libraries that satisfy
OpenEXR symbols need to appear after the OpenEXR archives. If your package
does not use `OpenEXRUtil`, you can omit `-lOpenEXRUtil-4_0`.

### C++ Example

The following `.Call()` example reads an EXR file into an R array with
dimensions `height x width x 4`.

```cpp
#define R_NO_REMAP

#include <R.h>
#include <Rinternals.h>

#include <OpenEXR/ImfChannelList.h>
#include <OpenEXR/ImfFrameBuffer.h>
#include <OpenEXR/ImfHeader.h>
#include <OpenEXR/ImfInputFile.h>
#include <Imath/ImathBox.h>

#include <cstddef>
#include <exception>
#include <vector>

using namespace OPENEXR_IMF_NAMESPACE;
using namespace IMATH_NAMESPACE;

extern "C" SEXP C_read_exr_array(SEXP path_sexp) {
  const char *path = CHAR(STRING_ELT(path_sexp, 0));

  try {
    InputFile file(path, 1);
    const Header &header = file.header();
    Box2i data_window = header.dataWindow();

    const int width = data_window.max.x - data_window.min.x + 1;
    const int height = data_window.max.y - data_window.min.y + 1;
    const int pixels = width * height;

    std::vector<float> red(pixels);
    std::vector<float> green(pixels);
    std::vector<float> blue(pixels);
    std::vector<float> alpha(pixels, 1.0f);

    FrameBuffer frame_buffer;
    const size_t x_stride = sizeof(float);
    const size_t y_stride = sizeof(float) * static_cast<size_t>(width);
    const ptrdiff_t offset =
      data_window.min.x * static_cast<ptrdiff_t>(x_stride) +
      data_window.min.y * static_cast<ptrdiff_t>(y_stride);

    frame_buffer.insert(
      "R",
      Slice(FLOAT, reinterpret_cast<char *>(red.data()) - offset, x_stride, y_stride)
    );
    frame_buffer.insert(
      "G",
      Slice(FLOAT, reinterpret_cast<char *>(green.data()) - offset, x_stride, y_stride)
    );
    frame_buffer.insert(
      "B",
      Slice(FLOAT, reinterpret_cast<char *>(blue.data()) - offset, x_stride, y_stride)
    );

    if (header.channels().findChannel("A") != nullptr) {
      frame_buffer.insert(
        "A",
        Slice(FLOAT, reinterpret_cast<char *>(alpha.data()) - offset, x_stride, y_stride)
      );
    }

    file.setFrameBuffer(frame_buffer);
    file.readPixels(data_window.min.y, data_window.max.y);

    SEXP out = PROTECT(Rf_allocVector(REALSXP, pixels * 4));
    double *dst = REAL(out);

    for (int y = 0; y < height; ++y) {
      for (int x = 0; x < width; ++x) {
        const int row_major = y * width + x;
        const int r_index = y + x * height;

        dst[r_index] = red[row_major];
        dst[r_index + pixels] = green[row_major];
        dst[r_index + pixels * 2] = blue[row_major];
        dst[r_index + pixels * 3] = alpha[row_major];
      }
    }

    SEXP dim = PROTECT(Rf_allocVector(INTSXP, 3));
    INTEGER(dim)[0] = height;
    INTEGER(dim)[1] = width;
    INTEGER(dim)[2] = 4;
    Rf_setAttrib(out, R_DimSymbol, dim);

    UNPROTECT(2);
    return out;
  } catch (const std::exception &err) {
    Rf_error("OpenEXR read error: %s", err.what());
  }

  return R_NilValue;
}
```
